Performance-portable Parallel Programming in Computer Science Education with Kokkos

### 2020-01-24 src/1.Introduction.tex
Changes after review with Christian

Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS}, is a performance-portable, parallel programming model for C++ developed at Sandia National Laboratories (SNL). It consists of a programming model specification and a library. The specification defines execution and memory model and an application programming interface (library API). The runtime library exposes programming primitives, namely, patterns and types and maps them to the underlying abstract machine model. The machine model creates a generic view of compute hardware and memories. The Kokkos library maps execution primitives to vendor-specific programming models, libraries, and memory layouts. Kokkos has been successfully used in many large-scale projects at SNL, LLNL and other research centers~\cite{KOKKOSUSECASE2}. We list its key properties as follows.

\begin{itemize}
\item Parallel programming model for C++
\item Targets an abstract machine model
\item Implements parallel patterns and tasking paradigms
\item Offers abstract data types for performance portability
\item Pure C++ implementation through template metaprogramming
\item Aligns with C++ standardization efforts for parallel programming
\item Vendor- and architecture independent and open-source
\item Includes exercises and teaching material
\end{itemize}

The objective of this paper is to present Kokkos from two perspectives. Firstly, we would like to present our thoughts that accompanied the design and development of the Kokkos parallel programming model. These insights may be considered useful in parallel programming classes that introduce programming models and reason about design choices.

Further, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards portability abstractions. We show a set of types that implement patterns and encapsulate data with enough semantic information to produce performance-portable, parallel code on current architectures. Further, we highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++.

Secondly, we would like to promote Kokkos to students and educators as a platform that helps them to understand and experiment with parallel patterns in a fashion that is generic, vendor-agnostic and declarative. This means that students can learn about parallelism without being exposed to many architectural concerns and that they can see productivity benefits of such abstractions for performance portability when programming on current and upcoming hardware architectures.

The rest of this paper is structured as follows. The next Chapter gives an overview on parallel programming concepts and presents the notion of \emph{semantic capture}. Chapter~\ref{chap:kokkosMM} presents the abstract machine model used to derive required semantic information for performance portability in Kokkos. Chapters~\ref{chap:kokkosPM} and ~\ref{chap:kokkosBackend} present the Kokkos programming model and its back-end support. Chapters~\ref{chap:c++} and~\ref{chap:related} show the alignment to C++ standardization efforts and dicuss related work and lastly Chapter~\ref{chap:conclusion} concludes this work.

>>>>

Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS}, is a performance-portable, parallel programming model for C++ developed through a multi-institutional collaboration between US national laboratories as part of the US Department of Energy Exascale Project\cite{ECP}. It consists of a programming model specification and a library. The specification defines execution and memory model and an application programming interface (library API). The runtime library exposes programming primitives, namely, patterns and types and maps them to the underlying abstract machine model. The machine model creates a generic view of compute hardware and memories. The Kokkos library maps execution primitives to vendor-specific programming models, libraries, and memory layouts. Kokkos has been successfully used in many large-scale projects at SNL, Utah University, ORNL, LANL and other research centers~\cite{KOKKOSUSECASE2}. We list its key properties as follows.

\begin{itemize}
\item Parallel programming model for C++
\item Targets an abstract machine model
\item Implements parallel patterns and tasking paradigms
\item Offers abstract data types for performance portability
\item Pure C++ interface through template metaprogramming
\item Aligns with C++ standardization efforts for parallel programming
\item Vendor- and architecture independent and open-source
\item Includes exercises and teaching material
\end{itemize}

The objective of this paper is to present Kokkos from two perspectives. Firstly, we would like to present our thoughts that accompanied the design and development of the Kokkos parallel programming model. These insights may be  useful in parallel programming classes that introduce programming models and reason about design choices.

Further, we would like to promote Kokkos to students and educators as a platform that helps them to understand and experiment with parallel patterns in a fashion that is generic, vendor-agnostic and declarative. This means that students can learn about parallelism without being exposed to many architectural concerns and that they can see productivity benefits of such abstractions for performance portability when programming on current and upcoming hardware architectures.

Finally, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards portability abstractions. We show a set of types that implement patterns and encapsulate data with enough semantic information to produce performance-portable, parallel code on current architectures. Further, we highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++.

The rest of this paper is structured as follows. The next Chapter gives an overview on parallel programming concepts and presents the notion of \emph{semantic capture}. Chapter~\ref{chap:kokkosMM} presents the abstract machine model used to derive required semantic information for performance portability in Kokkos. Chapters~\ref{chap:kokkosPM} and ~\ref{chap:kokkosBackend} present the Kokkos programming model and its back-end support. Chapters~\ref{chap:c++} and~\ref{chap:related} show the alignment to C++ standardization efforts and discuss related work and lastly Chapter~\ref{chap:conclusion} concludes this work.

### 2020-01-24 src/2.GenericProgramming.tex
Changes after review with Christian

\emph{Constrained semantics} provided by the programmer to the parallel programming model can be grouped into information to express intent or correctness and properties that control critical performance aspects. They address the question of \emph{what}, \emph{where} and \emph{how}. In the context of parallel programming, these correspond to defining which code portion to parallelize, where to run the code, where to access the data, and how to run that parallel code. Synchronization primitives may be considered part of the \emph{what}, and information on the execution properties such as data placement or memory access type may be considered as part of the \emph{how}. Application logic is expressed following a \emph{paradigm} implemented in a \emph{programming language}.

A parallel programming paradigm is an abstract representation with the purpose of facilitating the programmer's understanding of programming rules and program behavior. Which programming paradigm to choose depends on several considerations. Parallel patterns allow the programmer to permit concurrent execution for commonly occurring programming patterns such as loop constructs. Tasking is a paradigm that supports the expression of interdependent work items needed for irregular algorithms. Distributed and correctness-oriented programming models may implement actor-based programming, where each unit of execution represents an actor who communicates over predefined communication channels. This paradigm eliminates accesses to shared state and aligns well with message passing programming (MPI). The execution model and memory model aspects of a programming model define its behavior. That is, it defines the relationship between abstract concepts and program execution on the given architecture.

@@

Lastly, a programming model must divide responsibility between the developer and the model. We refer to is as defining the concern. In practice, this is a choice between defining a descriptive and prescriptive semantics. Descriptive semantics require a developer to express intent. The implementation is the responsibility of the programming model and of the underlying toolchain. Prescriptive semantics require the developer to specify the mechanisms by which their intent will be implemented. In an prescriptive model, developer success is tied to their knowledge of implementation mechanisms. In a descriptive model that success is tied to the developer's ability to accurately describe their requirements and the toolchain's ability to turn that into an implementation.

@@

Both examples provide the semantic information on what to parallelize, however, they do not expose enough abstraction primitives that would allow to map parallel execution on modern architectures. In order to define what further information is needed to express concurrency on modern computer architectures, it is important to define an~\emph{abstract machine representation} first. In the following chapters we use the words \emph{representation} and \emph{model} interchangeably. In this work the abstract machine model does not describe a computer in terms of automata theory but in terms of physical or functional components.

>>>>

\emph{Constrained Semantics} provided by the programmer to the parallel programming model can be grouped into information to express intent or correctness and properties that control critical performance aspects. They address the question of \emph{what}, \emph{where} and \emph{how}. In the context of parallel programming, these correspond to defining which code portion to parallelize, where to run the code, where to access the data, and how to run that parallel code. Synchronization primitives may be considered part of the \emph{what}, and information on the execution properties such as data placement or memory access type may be considered as part of the \emph{how}. Application logic is expressed following a \emph{paradigm} implemented in a \emph{programming language}.

A \emph{Parallel Paradigm} is an abstract representation with the purpose of facilitating the programmer's understanding of programming rules and program behavior. Which programming paradigm to choose depends on several considerations. Parallel patterns allow the programmer to permit concurrent execution for commonly occurring programming patterns such as loop constructs. Tasking is a paradigm that supports the expression of interdependent work items needed for irregular algorithms. Distributed and correctness-oriented programming models may implement actor-based programming, where each unit of execution represents an actor who communicates over predefined communication channels. This paradigm eliminates accesses to shared state and aligns well with message passing programming (MPI). The execution model and memory model aspects of a programming model define its behavior. That is, it defines the relationship between abstract concepts and program execution on the given architecture.

@@

Lastly, a programming model must divide responsibility between the developer and the model. We refer to is as defining the \emph{Control Paradigm}. In practice, this is a choice between defining a descriptive and prescriptive semantics. Descriptive semantics require a developer to express intent. The implementation is the responsibility of the programming model and of the underlying toolchain. Prescriptive semantics require the developer to specify the mechanisms by which their intent will be implemented. In a prescriptive model, developer success is tied to their knowledge of implementation mechanisms. In a descriptive model that success is tied to the developer's ability to accurately describe their requirements and the toolchain's ability to turn that into an implementation.

@@

Both examples provide the semantic information on what to parallelize, however, they do not expose enough abstraction primitives that would allow to map parallel execution on modern architectures efficiently. In order to define what further information is needed to express concurrency on modern computer architectures, it is important to define an~\emph{abstract machine representation} first. In the following chapters we use the words \emph{representation} and \emph{model} interchangeably. In this work the abstract machine model does not describe a computer in terms of automata theory but in terms of physical or functional components.

### 2020-01-24 src/4.ProgammingModel.tex
Changes after review with Christian

Kokkos defines a pure interface for C++ and uses C++ as a programming language. This has multiple reasons. It is our experience that application developers of scientific and high-performance computing applications at Sandia National Laboratories express the demand for C++ and a generic parallel programming model to support their codes. Following their estimates, porting efforts towards adapting to vendors, programming models, APIs and software releases represents a burden worth addressing. Further, the possibility to maintain pure C++ codes is appealing from the code development and debugging perspective. 
Lastly, from the programming model perspective, C++ offers template metaprogramming which is well suited to implement generic APIs and libraries. In this case, class specialization and templating allow for compile-type generated types and optimizations for a given hardware architecture. It is worth to mention that Kokkos avoids the use of pragma annotations for the sake of treating concurrency as a first-class concept in C++.

Using the programming language, a parallel programming paradigm can be applied to simplify the process of thought. Kokkos supports the parallel patterns and task paradigms. Parallel patters covers \emph{for}, \emph{ scan} and \emph{reduction}. This allows the expression of concurrency over iterative, for-loop computable algorithms. To cover the class of while-loop computable algorithms (irregular algorithms), Kokkos implements the tasking paradigm. Tasks encapsulate work into units that may be executed in parallel to other tasks or sections of the program. Equipped with a language and set of paradigms, a particular set of abstractions can be defined to expose the abstract machine model and captures the intent of the programmer.

To capture semantic information, that is the~\emph{what}, the~\emph{where} and the~\emph{how}, Kokkos introduces six abstractions:~\emph{execution spaces},~\emph{execution patterns},~\emph{execution policies},~\emph{memory spaces},~\emph{memory layout} and~\emph{memory traits}. These abstractions specify semantic information that enable to capture the programmer's intent and allow the runtime to efficiently map the program to any underlying hardware architecture. We list them as follows:
\begin{itemize}
	\item  An execution space is a place where code can be executed. On current hardware architectures this correspond to accelerators and CPUs and can include any compute device in the future. This abstraction supports remote compute devices in distributed memory scenarios as remote execution spaces.
	\item Execution patterns expose the parallel programming paradigm. Supported patterns are \emph{parallel\_for} loop that executes the loop body in any order a specified amount of times, the \emph{parallel\_reduce} which combines a parallel\_for with a reduction operation, \emph{parallel\_scan} which combines a parallel\_for operation with a prefix or postfix scan, and \emph{task} which executes a single function potentially in parallel in respect to other tasks or code sections. 
	\item Execution policy shape the iteration space of a loop pattern. A simple execution policy is a range policy. It specifies that the loop body is executed once for each element in a range.
	\item Memory spaces are the places where data resides. They specify physical locations of data as well as access characteristics. Different physical locations correspond to different device types such as high bandwidth memories, on-die scratch memories or non-volatile bulk storage. Different logical memory spaces allow for concepts such as memory in the CUDA programming model, which is accessible from the host and the CUDA accelerator. Memory spaces, similarly to execution spaces, conceptually support remote memory locations in distributed-memory scenarios. Furthermore, they encapsulate functionality such as consistency control and persistence scopes.
	\item Layouts express the mapping from loop indices to address offsets. By adopting appropriate layouts for memory structures, an application can optimize data access patterns in a given algorithm. If an implementation provides polymorphic layouts (i.e. a data structure can be instantiated at compile or runtime with different layouts), architecture-dependent optimizations can be performed.
	\item Memory traits specify how a data structure is accessed. Traits express usage scenarios such as atomic access, random access and streaming loads or stores. This allows the programming model to  optimize load and store operations.
\end{itemize}

@@

Figure~\ref{fig:parallelFor} shows one definition of the Kokkos~\emph{parallel\_for} interface. This interface accepts two template parameters and three function arguments. \emph{ExecPolicy} shapes the iteration space for a particular execution space. Valid template parameter are the~\emph{RangePolicy},~\emph{TeamPolicy},~\emph{MDRangePolicy},~\emph{TeamThreadRange} and~\emph{ThreadVectorRange}. As the name suggests, these policies express how an iteration space is mapped to the underlying hardware. For example, a range policy defines a one dimensional iteration range. A team policy defines a 1D iteration range, each of which is assigned to a thread team. This policy allows the expression of hierarchical parallelism. The second template parameter,~\emph{FuntorType}, represent a functor that implements the~\emph{()-operator} with a matching signature for the given execution policy. The functor can be defined using a C++ class, struct or lambda expression. We would like to highlight that the return semantic of the parallel\_for is defined as \emph{potentially asynchronous}. To guarantee a kernel has finished, a developer should call the fence of the execution space on which the kernel is being executed. Otherwise it depends on the execution space where the loop executes and weather this execution space implements a barrier. The semantic of the parallel\_for class in Kokkos demonstrates the descriptive nature of the programming model. To the interested reader, the complete programming model specification can be accessed on-line\cite{pub:KOKKOS}. Figure~\ref{fig:abstractions} summarizes patterns and abstractions and structures them by information type in the semantic capture.

Programming with patterns and abstractions allows the compiler to generate optimized transformations. In case of parallel patterns, the execution is unspecified and only promises deterministic results for the reduction and scan operations. This gives the freedom to the underlying abstraction layers to implement different mapping patterns on different hardware such as assignment of iterations to threads or vector lanes. We believe that these abstraction and their descriptive semantic represent key programming primitives and the way forward for generic support for parallel programming. In conclusion, Figure~\ref{fig:SemCaptureKokkos} shows the semantic capture defined by the Kokkos programming model. 
%Figure~\ref{fig:stack} shows abstraction layers in Kokkos and and how it fits in a parallel programming landscape.
The next section shows how the Kokkos parallel\_for class is mapped to the the underlying back-end programming model.

>>>>

Kokkos defines a pure interface for C++ and uses C++ as a programming language. This has multiple reasons. It is our experience that application developers of scientific and high-performance computing applications in the high-performance computing community and in industry in general express the demand for C++ and a generic parallel programming model to support their codes. Further, following their estimates, porting efforts towards adapting to vendors, programming models, APIs and software releases represents a burden worth addressing. Lastly, the possibility to maintain pure C++ codes is appealing from the code development and debugging perspective.

From the programming model perspective, C++ offers template metaprogramming which is well suited to implement generic APIs and libraries. In this case, class specialization and templating allow for compile-type generated types and optimizations for a given hardware architecture.

Using the programming language, a parallel programming paradigm can be applied to simplify the process of thought. Kokkos supports parallel patterns and tasking. Parallel patterns covers \emph{for}, \emph{scan} and \emph{reduction}. This allows the expression of concurrency over iterative, for-loop computable algorithms. To cover the class of while-loop computable algorithms (irregular algorithms), Kokkos implements the tasking paradigm. Tasks encapsulate work into units that may be executed in parallel to other tasks or sections of the program. Equipped with a language and set of paradigms, a particular set of abstractions can be defined to expose the abstract machine model and capture the intent of the programmer.

To capture semantic information, that is the~\emph{what}, the~\emph{where} and the~\emph{how}, Kokkos introduces six abstractions:~\emph{execution spaces},~\emph{execution patterns},~\emph{execution policies},~\emph{memory spaces},~\emph{memory layout} and~\emph{memory traits}. These abstractions specify semantic information that enable to capture the programmer's intent and allow the runtime to efficiently map the program to any underlying hardware architecture. We list them as follows:
\begin{itemize}
	\item  An execution space is a place where code can be executed. On current hardware architectures this correspond to accelerators and CPUs and can include any compute device in the future. This abstraction supports remote compute devices in distributed memory scenarios as remote execution spaces.
	\item Execution patterns expose the parallel programming paradigm. Supported patterns are \emph{parallel\_for} loop that executes the loop body in any order a specified amount of times, the \emph{parallel\_reduce} which combines a parallel\_for with a reduction operation, \emph{parallel\_scan} which combines a parallel\_for operation with a prefix or postfix scan, and \emph{task} which executes a single function potentially in parallel in respect to other tasks or code sections. 
	\item Execution policy shape the iteration space of a loop pattern. A simple execution policy is a range policy. It specifies that the loop body is executed once for each element in a range.
	\item Memory spaces are the places where data resides. They specify physical locations of data as well as access characteristics. Different physical locations correspond to different device types such as high bandwidth memories, on-die scratch memories or non-volatile bulk storage. Different logical memory spaces allow for concepts such as memory in the CUDA programming model, which is accessible from the host and the CUDA accelerator. Memory spaces, similarly to execution spaces, conceptually support remote memory locations in distributed-memory scenarios. Furthermore, they encapsulate functionality such as consistency control and persistence scopes.
	\item Layouts express the mapping from array indices to address offsets. By adopting appropriate layouts for memory structures, an application can optimize data access patterns in a given algorithm. If an implementation provides polymorphic layouts (i.e. a data structure can be instantiated at compile or runtime with different layouts), architecture-dependent optimizations can be performed.
	\item Memory traits specify how a data structure is accessed. Traits express usage scenarios such as atomic access, random access and streaming loads or stores. This allows the programming model to  optimize load and store operations.
\end{itemize}

@@

Figure~\ref{fig:parallelFor} shows one definition of the Kokkos~\emph{parallel\_for} interface. This interface accepts two template parameters and three function arguments. \emph{ExecPolicy} shapes the iteration space for a particular execution space. Examples of execution policies are the~\emph{RangePolicy}, ~\emph{MDRangePolicy} and the~\emph{TeamPolicy}. As the name suggests, these policies express iteration spaces. For example, a range policy defines a one dimensional iteration range. The MDRangePolicy represents a multi-dimensional iteration space and is used to express tightly nested loop patterns. A team policy defines a 1D iteration range, each of which is assigned to a team of threads. This policy allows the expression of hierarchical parallelism. The second template parameter,~\emph{FuntorType}, represent a functor that implements the~\emph{()-operator} with a matching signature for the given execution policy. The functor can be defined using a C++ class, struct or lambda expression. We would like to highlight that the return semantic of the parallel\_for is defined as \emph{potentially asynchronous}. To guarantee a kernel has finished, a developer should call the fence of the execution space on which the kernel is being executed. Otherwise it depends on the execution space where the loop executes and weather this execution space implements a barrier.

The semantic of the parallel\_for class in Kokkos demonstrates the descriptive nature of the programming model. Programming with patterns and abstractions allows the compiler to generate optimized transformations. 
Generally parallel patterns in Kokkos do not guarantee iteration ordering nor degree of concurrency during their execution. This gives the freedom to the underlying abstraction layers to implement different mapping patterns on different hardware such as assignment of iterations to threads or vector lanes. We believe that these abstraction and their descriptive semantic represent key programming primitives and the way forward for generic support for parallel programming. In conclusion, Figure~\ref{fig:SemCaptureKokkos} shows the semantic capture defined by the Kokkos programming model. 
%Figure~\ref{fig:stack} shows abstraction layers in Kokkos and and how it fits in a parallel programming landscape.
To the interested reader, the complete programming model specification can be accessed on-line\cite{pub:KOKKOS}. Figure~\ref{fig:abstractions} summarizes patterns and abstractions and structures them by information type in the semantic capture.
The next section shows an example of the Kokkos parallel\_for and explains how its mapped to the the underlying back-end programming model.

### 2020-01-24 src/8.Related.tex
Fused DZP and JC changes for Chapters 1 and 2.
Changes after review with Christian

\section{Related Work}\label{chap:related}
Like Kokkos, RAJA~\cite{RAJA} is a C++ framework for porting massive DOE codes to a variety of architectures. In both frameworks, a developer first expresses the operations their kernel performs using simple C++ functional programming mechanisms. The difference is that Kokkos is declarative, while RAJA is imperative. In Kokkos, the framework figures out how to map that kernel to different architectures. In RAJA, functionality is provided to make it easy for a developer to try different strategies to map those kernels to architectures, but the developer still decides how that mapping happens. Which model makes more sense in a given circumstance depends on how much exposure to the underlying programming model a developer wants. RAJA is a viable option for class education in which the educator wants to expose students to the semantics of different programming models without exposing them to the detailed syntax of those models.

DPC++\cite{DPCPP} is a model developed by Intel\textsuperscript{\textregistered} for expressing parallelism across Intel architectures. In addition to sharing a declarative philosophy with Kokkos, it also shares parallel patterns. Unlike Kokkos, at time of publication it targets only Intel architectures and is not vendor-agnostic. A course targeting Intel architectures, DPC++ could be a valid option.

C++ increasingly supports parallel constructs\cite{CPP}. We have highlighted contributions of Kokkos to C++ in~\ref{chap:c++}. It is becoming increasingly possible to teach parallelism in C++ and relying on modern C++ compilers. However, at time of publication, C++ does not support accelerators, or many of the complexities of parallel computing related to accelerator programming. In courses that do not target accelerators, this can be a relevant option.
%\end{itemize}

>>>>

\section{Related Work}\label{chap:related}
Like Kokkos, RAJA~\cite{RAJA} is a C++ framework for porting massive DOE codes to a variety of architectures. In both frameworks, a developer first expresses the operations their kernel performs using simple C++ functional programming mechanisms. The difference is that Kokkos is declarative, while RAJA is descriptive. In Kokkos, the framework figures out how to map that kernel to different architectures. In RAJA, functionality is provided to make it easy for a developer to try different strategies to map those kernels to architectures, but the developer still decides how that mapping happens. Which model makes more sense in a given circumstance depends on how much exposure to the underlying programming model a developer wants. RAJA is a viable option for class education in which the educator wants to expose students to the semantics of different programming models without exposing them to the detailed syntax of those models.

DPC++\cite{DPCPP} is a model developed by Intel\textsuperscript{\textregistered} for expressing parallelism across Intel architectures. In addition to sharing a descriptive philosophy with Kokkos, it also shares parallel patterns. Unlike Kokkos, at time of publication it targets only Intel architectures and is not vendor-agnostic. A course targeting Intel architectures, DPC++ could be a valid option.

C++ increasingly supports parallel constructs\cite{CPP}. We have highlighted contributions of Kokkos to C++ in~\ref{chap:c++}. It is becoming increasingly possible to teach parallelism in C++ and relying on modern C++ compilers. However, at time of publication, C++ does not support accelerators, or many of the complexities of parallel computing related to accelerator programming. In courses that do not target accelerators, this can be a relevant option.

% Alpaka Dresden
% NV Thrust
